PanGraph: scalable bacterial pan-genome graph construction

### 2023-03-08 reviewer_response.tex

      There was no comparison with other tools - would be appropriate if e.g. downstream mapping performance were evaluated but that is outside the scope here.\\

      On reading I had a few queries, which could perhaps be clarified further in the text.}

\response{We thank the reviewer for their appreciation of our contribution and for their critical review of our work.}

\reviewer{R1.1}{A pancontig is defined as a multiple sequence (sub)alignment. In line 54 you say "Pairwise graph alignment is performed by an all-to-all alignment of the pancontigs" which I interpreted to mean all sequences in the alignment vs all sequences in another alignment. You do later add that ``To align two graphs, the consensus sequences of all pancontigs in both graphs are searched for homologies and aligned". Adding the word consensus in the earlier sentence might avoid confusion.}
% \Marco{changed to ``Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree." Too compressed? I could also break in two sentences and expand a bit more.}\\
% \Liam{that's clear to me.}
\response{We agree with the reviewer's suggestion. The wording has been updated to ``all-to-all alignment of the consensus sequences of all \textit{pancontigs}''.}

@@

\response{The typos have been corrected.}

\reviewer{R1.4}{Throughout the text E. Coli and K. Pneumoniae etc capitalization.}

\response{The capitalization has been corrected.}

\reviewer{R1.5}{Line 213: Core pancontigs have higher than average size - is this what you would expect given that if they are core genes then they will be present in more/all of the samples so there could be more diversity due to being more sampled?}
% \Marco{here the pancontig size is the consensus sequence length. I would say that the fact that they have higher-than-average size is mostly due to synteny. Long blocks (several kbps) contain core genes that always appear in the same order, without any accessory sequence ever occurring in these regions. As opposed to accessory regions with a lot of structural diversity, that generates short fragments. But here the main message is that core blocks are well-behaved (close to N50 line in the figure), as opposed to fine-grained structural diversity, part of which contains artifacts generated by the block size limit (many short 100bp blocks). Should we show block frequency distributions in the supplementary?}\\
% \Liam{I don't understand their point on diversity - maybe they misunderstood the meaning of size. Maybe pancontig `length' is clearer? I agree with you about synteny and think this is a nice point to add here to explain the observation.}
\response{The observation that core pancontigs are longer than average is in agreement with our expectations. Multiple core genes are often syntenic even in diverse species, compared to accessory genes that are often found in different contexts. This effect seems to outweigh the possibility of increased fragmentation due to the large number of mergers that common (core) pancontigs undergo.
      We modified the line to: ``Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay syntenic over large evolutionary distances.''}

@@

\reviewer{R1.6}{Figure 6B: could you explore another way to project this data so it is not squished at the ends? I can't read off the fraction agree (shared+private) or disagree which I would like to (although at a glance this is exactly the result you want to see, so happy for it to stand if it is the best projection).}
% \Marco{maybe add mean and standard deviation as numbers on the side?}\\
% \Liam{that seems ideal.}
\response{We considered different projections and in the end decided to maintain the linear projection for ease of interpretation, but we added next to each distribution its mean and standard deviation. In this way both deviations from 0 and 1 in the first two distributions can be quantified.}

\reviewer{R1.7}{Have you tried running PanGraph on more than 300 references? I'd love to see another order of magnitude (not necessary for the paper, just curiosity)}
% \Marco{not really, I never tried on ~500 or ~1000 strains. I could try but would not consider this a priority.}\\
% \Liam{agree that it's not necessary.}
\response{The biggest dataset we tried PanGraph on is a collection of 500 E.coli isolates. We used minimap2 as alignment kernel, asm20 sensitivity and values of $\alpha = 100$, $\beta=20$. Running on 8 cores this took 18 hours and 12 minutes to complete and had maximum memory usage of 17Gb.}

\reviewer{R1.8}{Discussion: Do you know of/have you tried any mapping approaches which could take this reference as an input. Could you touch on the downstream applications of this graph structure? I would like to see discussion of how the graph can be used downstream - e.g. touching on what software can be used to map/align to such a data structure? Or how it can be used to analyze genome structure.}
% \Marco{other than explicitly exporting block sequences and mapping to those, we do not really using the pangenome graph as a reference against which to map new strains... But we could expand the discussion with more ideas on applications to study pangenome organization and evolution. For software for further analysis and exploration I have \href{https://github.com/mmolari/pypangraph}{this package} to explore and interact with graphs in python, but not sure it's worth mentioning.}\\
% \Liam{Some ideas that could be mentioned. (1) I think one obvious (bad!) use would be to take the core blocks and map reads to those as a `core reference'. (2) I know pandora has \href{https://github.com/iqbal-lab-org/make_prg/}{\texttt{make\_prg}} which will construct a `pangenome reference graph' from a set of aligned sequences - I didn't try it on pangraph output. (3) For genome structure, exploring `hotspots' in the graph as \href{https://www.nature.com/articles/s41467-017-00808-w}{this paper} does could be done using in/out degree of pancontigs.}
\response{We have so far not explored the use of pangraph as input for graph based mapping approaches like Pandora or GramTools, but we agree this is an interesting downstream application. We have included a discussion of these approaches, although we feel that testing and implementation of them is beyond the scope of this article.}

\reviewer{R1.9}{Supplementary Information: simulation - could you comment on how/if homologous recombination is included in this model e.g. within a gene.}
% \Marco{homologous recombination is not implemented directly in the model. Only sequence transfer between strains. Genes are not explicitly modeled in the simulation.}\\
% \Liam{think just adding a comment in the SI saying this is required.}
\response{Homologous recombination is not directly included in our simulated data generation procedure. We only simulate transfer of sequence between two genomes but this transfer is not based on homology. Moreover we simulate sequence evolution without a notion of genes. This is now specified in SI section \SIsynthdata. To avoid confusion we also changed ``Horizontal Gene Transfer'' to ``Horizontal Sequence Transfer''.}

\reviewer{R1.10}{Supplementary Information C: You comment that the use of null energy parameters results on average in ``more and shorter pancontigs" but also that they result in more diverged sequences being merged into the same pancontig. Why is a more diverse pancontig more likely to become fragmented?}
% \Marco{I think that basically the reason is that merging more distant paralogs/orthologs can fragment more the graph. Maybe we could explain this more explicitly in the text.}\\
% \Liam{makes sense and explaining this reasoning sounds good.}
\response{We use the pseudo-energy defined in eq.~(1) as an heuristic to decide whether a homologous match detected by our aligner should result in a merge between two pancontigs, depending on the match length and on sequence divergence. In particular any choice of $\beta$ would prevent mergers between sequences with divergence higher than $1/\beta$. Setting $\beta=0$ instead does not impose a divergence limit on mergers other than the one implicitly set by the aligner sensitivity. In this sense the choice of null energy parameter results in more diverged sequences being merged into the same pancontig. A consequence of more merging is also more fragmentation of the graph, since the merging of a homologous section of two pancontigs often requires splitting them in chunks. Therefore we expected that the choice of null energy parameters would lead to shorter pancontigs, and alignments containing more diverged sequences.
      To make this reasoning more explicit we added section {\SIalgo} in the SI to better explain the merging procedure and the effect of tuning the pseudo-energy parameters.}

@@

      Nevertheless, there are a number of clarifications which need to be addressed before the paper is suitable for publication.}

\response{We thank the reviewer for this succinct summary of the field and the essence of our contribution.}

\subsubsection*{Major Issues}

\reviewer{R2.1}{Line 46: when you say ``PanGraph transforms an arbitrary set of genomes into a graph that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms." - I am a bit unclear on how the nucleotide level variation is summarised. The paper is a bit hand-wavey about this. When i dug around on github , i found this helpful page \url{https://neherlab.github.io/pangraph/tutorials/tutorial_2/} Which answers and raises questions. So, my question: for each pancontig (which I think is the same as a block), I think what you do is infer a consensus (what is this? majority base at each position but somehow also choose an allele when there is an indel?), and then you describe the other alleles in the MSA as edits from this? You surely do not do all pairwise descriptions of SNPS, right? On that same webpage you have a section on ``block consensus to node alignment'' - how do you choose the order of alleles here? This is all valuable information and it would be great to have it in the supplementary or main methods somewhere.}
% \Marco{this point was also raised by R3. Mutations are always represented as variation on the consensus. The way consensus is propagated during graph merging might require more explanation. Maybe we could add a section in the supplementary with a somewhat more detailed description of the algorithmic steps we perform when merging two graphs? In the code \href{https://github.com/neherlab/pangraph/blob/e1c11012a3d19975d4b1dc61b1af3eb3949f7cd2/src/block.jl\#L1274}{this part} takes care of merging alignments between two blocks with only small (i.e. below cutoff) insertions and deletions. In short what we do is stitch the two alignments together using the cigar string as a guide, and re-compute the consensus on the newly-obtained alignment. This ensures that the consensus agrees with the majority of sequences in the alignment, but does not solve the full multiple sequence alignment problem. However we also provide the ``polish'' step for this, that can be executed after graph creation. This will take care of refining the block alignments.}\\
% \Liam{I think a `narrative section' in the supplementary describing this in brief sounds great.}
\response{This point was also raised by other reviewers and we appreciate the need to make this point clearer.
      In our pangenome graph representation each pancontig encodes an alignment using a consensus sequence and list of edits required to obtain each individual sequence from the consensus sequence.
      These edits consist of single nucleotide changes (mutations) as well as insertions and deletions.
      The consensus sequence is defined using the majority rule on the alignment and is updated after each merger.

      To make this clearer we have added section {\SIgraph} in SI, in which we explain in more details our pangenome graph representation, and section {\SIalgo}B in SI, in which we explain in more detail the merging procedure and how the consensus sequence is updated. We also refer to the documentation page mentioned by the reviewer. Moreover, to avoid ambiguity we also avoided referring to pancontigs as ``blocks'' in the main text and use pancontigs throughout.}

\reviewer{R2.2}{Line 61: I agree and understand how each sequence can be sketched in linear time. I don't follow how you calculate all pairwise intersections through sorting the list of minimizers, nor do i see how this ends up being log-linear. For each genome, you can sort it's minimizers in $n \log (n)$ (ie log linear), where $n \sim$ genome length. How do you get all pairwise distances from that without another multiplier of $n$?}
% \Marco{in the text we write that ``the pairwise distance matrix is estimated in a time that increases log-linearly in the \textbf{total length} of sequence that is analyzed.'' This might generate some confusion since the scaling depends both on the number of genomes $N$ and their length $l$. The algorithmic time for all the pairwise distances should scale as $\sim N^2 l \log l$. We should be in a regime where $N$ is not a limiting factor, it's just a multiplicative constant, and thus I think our statement is fundamentally correct. But we can make it more precise. This scaling is at any rate not a limiting factor in pangraph.}\\
% \Liam{clarifying $N$ and $l$ makes sense.}
\response{We agree with the reviewer that the formulation in the main text was too concise and a bit inaccurate.
      There are two main variables at play: the length of a genome $l$ and the number of isolates $n$. As the reviewer points out, the set of minimizers for each genome can be created and sorted in $l \log l$. Finding the set of shared minimizers would require comparing $\sim n^2$ pairs of ordered lists of minimizers, which scales as $\sim n^2 l$. This would provide a scaling $O(n l \log l + n^2 l)$.\\
      In our implementation we achieve a slightly better scaling as a function of $n$ by sketching and sorting all the genomes together, which can be done in time $L \log L$, where $L= n \times l$ is the total length of all genomes. We keep the information on each minimizer origin in the list. We then scroll the list and for each chunk of equal minimizers we update the corresponding entries of a $n^2$ matrix of minimizer hit counts. Depending on the amount of shared minimizers, this scales as $n^2 l$ in the worst case (all minimizers shared) and as $n l$ in the best case (no minimizer shared). Therefore depending on the diversity of the sample the scaling will be between $O(L \log L + n L)$ and $O(L \log L)$.\\
      In practice, for typical use cases of pangraph ($n \sim 100$ and $L \sim 10^6$) the computationally challenging part is the minimizer sketching, and not the distance matrix computation. This is partly due to some prefactors (e.g. the window size of 100 bp used for sketching makes so that the list of minimizer is significantly shorter than the genome, and the only operation performed $n^2$ times is a single increment of an integer in the minimizer hit matrix). We also stress that the computational time to build the guide tree is usually negligible compared to the time required for the graph merging procedure.\\
      We slightly elaborated this point in the main text and added a version of the above explanation to SI sect. {\SIalgo}A.
}

\reviewer{R2.3}{The section on iterative graph alignment (lines 69-77) is in one sense quite clear and intuitive, but there is a lot of detail missing which leaves the reader a bit confused when they try to think about details. I would appreciate some detail being added, perhaps in the supplementary methods. For example:
      \begin{itemize}
            \item at the start you just have two genomes, and you scan them for orthologous regions, and you decide which to merge. As i understand it, you merge them in order of pseudoenergy ranking, which makes sense (what do you do for precise duplicates, where a region on genome 1 is a precise match for 2 regions on genome 2? random choice?)
                  % \Marco{the ranking would be the same and we perform only one of the two merges in the first phase, when comparing the two graphs. After this we align the graph on itself and the second merge should be performed at this point. This comment also suggests we could add a section in the supplementary with a more detailed description of the algorithm?}
            \item as you move on, you are aligning graphs, and you really skim over the issue of how you do that. How does minimap or mmseqs allow you to look for homology between two graphs? The only thing i can imagine you are doing is only looking for homology between the genomes in one graph, and the genomes in the second graph. I would really appreciate it if you could clarify this and even ideally add some example figures in the supplement showing with some toy genomes how you align two graphs.
                  % \Marco{another comment pointing at the fact that we could add a more detailed explanation about graph merging on the supplementary. What we do is using the aligner using as reference and queries the consensus sequences of each block in the two different graphs, and create a first version of the merged graph. We then repeat this procedure in an all-vs-all comparison of the consensus sequences of the blocks, and proceed with further merges. This is necessary because when two alignments overlap only one of the two is performed in the first alignment round, but the second will be performed in further rounds.}\\
                  % \Liam{agree, sounds like a section in supplementary is a good idea to address these comments.}
      \end{itemize}}
\response{We thank the reviewer for pointing out that we should have provided more explanation (the point was also raised by other reviewers).\\
      We followed the advice of the reviewer and added Section {\SIalgo}B in the SI, in which we expand on the explanation of how we merge two graphs.\\
      In short, we perform the merging procedure in two steps. In the first we use the aligner (minimap2 or mmseqs2) to scan for homology matches between the consensus sequences of all pancontigs of one graph and of the other.
      This is achieved by simply passing all these consensus sequences as a single fasta file into mmseqs2 or via the API into minimap2.
      The resulting matches are ranked by pseudo-energy and mergers performed in order.
      Each pancontig can be subject to only one merge at this stage.
      For exact ties in this ranking of of matches, the behavior is not explicitly defined but depends on the order in which the matches appear in the output of the homology search tool. % Liam: replaced 'precise ties' > 'exact ties'

      After all of the matches have been exhausted, the two graphs have undergone a preliminary merging, with some new merged pancontigs being created. At this point we repeat the procedure looking for matches between the consensus sequences of all the pancontigs in an all-against-all comparison. This self-mapping on the graph will detect any homology that could not have been merged at the previous stage, and will perform the corresponding merging. The self-mapping is repeated iteratively until no new merges can be performed.
      Exact duplicate matches would thus be merged in successive iterations of this merge-process.
}

\reviewer{R2.4}{Line 92-94, same section. You give two statements:
      ``Only mergers whose alignment has negative pseudo-energy are performed''
      and
      ``no mergers will be performed if the Hamming distance exceeds $1/\beta$.''
      These are separate criteria that prevent mergers, right? So any potential merger with negative pseudoenergy that does not fail these criteria will happen, with the caveat that you perform mergers in order?\\
      One interesting thing here is that you are effectively merging things in reverse evolutionary order, as you have lower energy for longer more similar things which presumably diverged more recently. Not sure if that tells us anything.}
% \Marco{the second is actually a consequence of the first, but except for that it is correct.}\\
% \Liam{maybe say `$\beta$ controls the maximal sequence divergence between the two merger-candidates. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.'}\\
% \Marco{indeed, the parameter $\beta$ could control how far in the past one wants to accept merges. Should we mention this explicilty in the text?}\\
% \Liam{could be mentioned as an `intuitively, this should correspond to merging of homologous stretches of sequence in roughly reverse evolutionary order'.}\\
\response{
      Apologies for these confusing statements.\\
      In this case the second criterion is a consequence of the first. From the definition of the pseudo-energy (cf. eq.~(1) in the main text) it follows that for an alignment of pancontig consensus sequences to have negative pseudo-energy must be that the number of SNPs divided by the alignment length $N_m/l$ must be less than $1/\beta$. Moreover, as the reviewer pointed out, the pseudo-energy ranking of merges effectively results in merges being performed in order of similarity, effectively in reverse evolutionary order. \\

      To make this more explicit we modified the previous formulation into: ``$\beta$ controls the maximal sequence divergence between the two merger-candidates. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$... The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.''
      % QUESTION: is there ambiguity in the main text on the value of N_m? As SNPs in the alignment? Misleading figure?
}

@@

\reviewer{R2.5}{I very much appreciated that you measure the accuracy of pangraph via this displacement idea. I started writing a kindly worded criticism that I thought your method was flawed - surely the displacement depends not only on the order of doing merges, and decisions made about what to do when there are two equally good places to merge with. I guess i was saying that slightly different evolutionary events could give the same genomes, and thereby justify the errors. However i very much like the approach you detail in the methods, numerically solving the assignment problem. So, this started as a major issue, but in the end is a compliment.}

\response{Thank you!}

### 2023-03-08 main.tex
Update on Overleaf.

\begin{abstract}
    The genomic diversity of microbes is commonly parameterized as single nucleotide polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
    However, any reference genome contains only a fraction of the microbial \emph{pangenome}, the \textit{total} set of genes observed in a given species.
    Reference-based approaches are thus blind to the dynamics of the accessory genome, as well as variation within gene order and copy number.
    With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
    % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
    This provides access to very rich information on the variation present in the population, both at the nucleotide level and as large-scale structural changes, that traditional computational approaches struggle to capture and represent.
    Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
    Each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
    The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
\end{abstract}

@@

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}, with tools existing to build graphs containing thousands of isolates \cite{holley2020bifrost}. These have proven useful in problems such as the detection of core pangenome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}. However in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements in the genome. Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.

An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
Some early methods relied on homology searches to locate syntenic regions, termed \textit{locally colinear blocks} \cite{angiuoli2011mugsy,darling2010progressivemauve}, and proved useful in various problems such as improving gene annotations \cite{angiuoli2011improving}. These early methods however usually scale poorly to large sets of genomes.
More recently techniques based on homology have been used to compare genomes across diverse sets of species across the tree of life, at the cost of memory \cite{armstrong2020progressive}.
Alternatively, homology can be searched directly on genes. Many methods rely first on grouping genes in orthologous clusters, representing nodes in the graph, with edges linking clusters that are adjacent on some of the genomes.
These graphs are usually characterized by a syntenic backbone of core gene clusters, with interspersed regions of structural diversity \cite{chan2015novel,oliveira2017chromosomal}.
The structural information provided in these graphs has been used to categorize core/accessory regions \cite{sutton2021pan,gautreau2020ppanggolin}, and to improve variants detection \cite{colquhoun2021pandora} or gene annotations \cite{tonkin2020producing,zhou2020accurate}. By using genes as nodes, these graphs are much smaller in size and homology detection is facilitated, but many of these methods are sensitive to gene annotation errors.

Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers. Alignment is performed on the base of sequence homology and does not rely on gene annotations.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss. Importantly, these two scales are naturally separated in our representation.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

@@

At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments. This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments. After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation).

@@

We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~\SIsynthdata for details on the simulation and standard parameter values).
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line tools for external use.

@@

\paragraph*{Accuracy as a function of sequence divergence}
To be useful and informative beyond data compression of microbial genomes, the contigs and break-points of the reconstructed graphs should correspond, at least approximately, to evolutionary events in the history of the sample.
We quantified PanGraph's ability to accurately reconstruct the true graph by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf. SI Section IIB for details). We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and horizontal sequence transfer $h$. For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel: \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} options and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal transfer $h$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf. SI fig.~\SIfigMutrate). For low-diversity datasets, breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf. SI fig.~\SIfigKernels). The choice of alignment kernel influences the threshold diversity at which accuracy is lost, suggesting that inability to detect homology between diverged sequences is the reason.\\
For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default minimal pancontig size for Pangraph of 100 bp (cf. Fig.~\ref{fig:toy-accuracy}). On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at an average pairwise sequence divergence of around 5\%, while \textit{mmseqs2} is accurate up to around 12\%, at the cost of higher computational time.
These threshold are lower than the nominal sensitivity of the aligners, which are 10\% and 20\% for minimap2 with settings \textit{asm10} and \textit{asm20}, respectively, and around 30\% for \textit{mmseqs2}.
This discrepancy is due to the fact that for accurate graph constructions, the largest divergence, rather than the average divergence, is the relevant quantity.

@@

Fig.~\ref{fig:panx-benchmark}B-C and SI fig.~\SIfigBenchmark quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
As expected, the very homogenous \textit{M.~tuberculosis} dataset has a large core genome with and a compression ratio that is almost equal to the inverse number of input genomes, independent of the aligner or parameters used.
\textit{E.~coli} and \textit{K.~pneumoniae} compress similarly well, but their core genome is a much smaller fraction of the pangenome.
Using the more sensitive aligner mmseqs2 and relaxing the merger parameters leads to some additional compression, presumably due to merging of diverged paralogs.

The \textit{H.~pylori}  dataset sits at the limit of what can be accurately merged with the \textit{minimap2} kernel.
In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf. Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~4).
The divergence of the \textit{P.~marinus} dataset sits instead beyond the capabilities of \textit{Pangraph}. In this case no alignment kernel can reach satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve a few core pancontigs.\\

>>>>

\begin{abstract}
    The genomic diversity of microbes is commonly parameterized as single nucleotide polymorphisms relative to a reference genome of a well-characterized, but arbitrary, isolate.
    However, any reference genome contains only a fraction of the microbial \emph{pangenome}, the \textit{total} set of genes observed in a given species.
    Reference-based approaches are thus blind to the dynamics of the accessory genome, as well as variation within gene order and copy number.
    With the wide-spread usage of long-read sequencing, the number of high-quality, complete genome assemblies has increased dramatically.
    % Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.
    In addition to pangenomic approaches that focus on the variation in the sets of genes present in different genomes, complete assemblies allow investigations of the evolution of genome structure and gene order.
    This latter problem, however, is computationally demanding and conceptually challenging with few tools available that allow to shed light on this dynamics. 
    Here, we present \emph{PanGraph}, a Julia-based library and command line interface for aligning whole genomes into a graph.
    Each genome is represented as an undirected path along vertices, which in turn, encapsulate homologous multiple sequence alignments.
    The resultant data structure succinctly summarizes population-level nucleotide and structural polymorphisms and can be exported into a several common formats for either downstream analysis or immediate visualization.
\end{abstract}

@@

While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging.
Colored and compacted generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets \cite{iqbal2012novo,muggli2017succinct}, with tools existing to build graphs containing thousands of isolates \cite{holley2020bifrost}. These graphs have been shown to be useful in problems such as the detection of the core genome \cite{schulz2022sequence} or the improvement of gene prediction, annotation and clustering \cite{horsfield2023accurate}. However, in these methods the graph structure encodes at the same time nucleotide-level variation and large-scale structural rearrangements. Separating these two scales is often difficult, and specific techniques have to be employed to detect large-scale homology in these graphs \cite{angiuoli2011mugsy,minkin2013sibelia}.

An orthogonal approach has been to formulate the inference of the pangenome graph as a multiple genome alignment.
Some early methods relied on homology searches to locate syntenic regions, termed \textit{locally colinear blocks} \cite{angiuoli2011mugsy,darling2010progressivemauve}, and proved useful in various problems such as improving gene annotations \cite{angiuoli2011improving}. These early methods however usually scale poorly to large sets of genomes.
More recently techniques based on homology have been used to compare genomes across diverse sets of species across the tree of life, at the cost of memory \cite{armstrong2020progressive}.
Alternatively, homology can be searched directly on genes. Many methods rely first on grouping genes into orthologous clusters. These clusters are then treated as nodes in the graph, with edges linking clusters that are adjacent on at least one of the genomes.
These graphs are usually characterized by a syntenic backbone of core gene clusters, with interspersed regions of structural diversity \cite{chan2015novel,oliveira2017chromosomal}.
The structural information contained in these graphs has been used to categorize core/accessory regions \cite{sutton2021pan,gautreau2020ppanggolin}, and to improve variant detection \cite{colquhoun2021pandora} or gene annotations \cite{tonkin2020producing,zhou2020accurate}. By using genes as nodes, these graphs are much smaller in size. However, relying on gene annotations means that these methods are often sensitive to any errors in this initial annotation.

Here we present \emph{PanGraph}, a Julia \cite{bezanson2017julia} library and command line interface, designed to efficiently align large sets of closely related genomes into a pangenome graph on personal computers. The \emph{PanGraph} algorithm creates nodes of homologous sequences through alignments alone, without requiring gene annotation.
The resulting graph both compresses the input sequence set and succinctly captures the population diversity at multiple scales: from nucleotide mutations and indels to structural polymorphisms driven by inversions, rearrangements, and gene gain/loss. Importantly, these two scales are naturally separated in our representation.
The underlying graph data structure can be exported into numerous formats for downstream analysis and visualization in software such as Bandage \cite{wick2015bandage}.
%The algorithm, along with extensive documentation, examples, and instructions for installation, are open source and can be found on GitHub.
%We anticipate PanGraph to be immediately useful in understanding microbial evolution in natural populations.

@@

At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments. This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments. After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation and SI sect.{\SIalgo}B).

@@

We simulated populations of size $N$ and sequence length $L$ utilizing a Wright-Fisher model \cite{hudson2002generating} evolved for $T=50$ generations.
In addition to nucleotide mutations that occur at rate $\mu$ per generation, we modelled inversions, deletions (respective rates $0.01$ and $0.05$ per generation), and horizontal sequence transfer that occurs with tunable rate $h$ per generation per genome (see SI sect.~{\SIsynthdata} for details on the simulation and standard parameter values).
The ancestral state for each sequence is tracked through each evolutionary event so that the true mosaic relatedness structure can easily be converted to a graph.
The simulation framework is distributed within the PanGraph command line tools for external use.

@@

\paragraph*{Accuracy as a function of sequence divergence}
To be useful and informative beyond data compression of microbial genomes, the contigs and break-points of the reconstructed graphs should correspond, at least approximately, to evolutionary events in the history of the sample.
We quantified PanGraph's ability to accurately reconstruct the true graph by computing the displacement of the inferred breakpoints relative to their known locus stored from the evolutionary simulation (cf. SI Section {\SIsimBench}B for details). We evaluate this displacement by generating datasets with different rates of mutation $\mu$ and horizontal sequence transfer $h$. For each $(h,\mu)$ pair we perform 25 different simulations, and build pangenome graphs using three different options for the alignment kernel: \textit{minimap2} \cite{li2018minimap2} with \textit{asm10} and \textit{asm20} options and \textit{mmseqs2} \cite{steinegger2017mmseqs2}.
Critically, we found that the accuracy was \emph{independent} of the rate of horizontal transfer $h$ and thus underlying graph complexity.
The predominant determinant of accuracy was the sample diversity, controlled by the mutation rate $\mu$ in our simulations (cf. SI fig.~\SIfigMutrate). For low-diversity datasets, breakpoints are inferred with accuracy of few basepairs, while for highly diverged isolates most breakpoints are displaced by several hundreds of basepairs (cf. SI fig.~\SIfigKernels). The choice of alignment kernel influences the threshold diversity at which accuracy is lost, suggesting that inability to detect homology between diverged sequences is the reason.\\
For each alignment kernel and value of average sequence divergence in the population we evaluated the fraction of breakpoints that have displacement greater than the default minimal pancontig size for Pangraph of 100 bp (cf. Fig.~\ref{fig:toy-accuracy}). On the sequences generated by our simulations, \textit{minimap2} with option \textit{asm20} show a loss of accuracy at an average pairwise sequence divergence of around 5\%, while \textit{mmseqs2} is accurate up to around 12\%, at the cost of higher computational time.
These threshold are lower than the nominal sensitivity of the aligners, which are 10\% and 20\% for minimap2 with settings \textit{asm10} and \textit{asm20}, respectively, and around 30\% for \textit{mmseqs2}.
This discrepancy is due to the fact that for accurate graph constructions, the largest divergence, rather than the average divergence, is the relevant quantity.

@@

Fig.~\ref{fig:panx-benchmark}B-C and SI fig.~{\SIfigBenchmark} quantify the size of the core genome, i.e.~the sum of all pancontigs present once in every genome, identified by PanGraph and the compression of the dataset, i.e.~the ratio for the sum of the lengths of all pancontigs and the sum of the lengths of all input genomes, for different dataset and the different parameters.
As expected, the very homogenous \textit{M.~tuberculosis} dataset has a large core genome with and a compression ratio that is almost equal to the inverse number of input genomes, independent of the aligner or parameters used.
\textit{E.~coli} and \textit{K.~pneumoniae} compress similarly well, but their core genome is a much smaller fraction of the pangenome.
Using the more sensitive aligner mmseqs2 and relaxing the merger parameters leads to some additional compression, presumably due to merging of diverged paralogs.

The \textit{H.~pylori}  dataset sits at the limit of what can be accurately merged with the \textit{minimap2} kernel.
In this case the use of \textit{mmseqs2} and null energy parameters increases core pangenome fraction, decreases the total pangenome size and and does not compromise the fragmentation of the graph (cf. Fig.~\ref{fig:panx-benchmark}B-C and SI Fig.~{\SIfigBenchmark}).
The divergence of the \textit{P.~marinus} dataset sits instead beyond the capabilities of \textit{Pangraph}. In this case no alignment kernel can reach satisfactory sequence compression, and only \textit{mmseqs2} combined with null energy parameters is able to retrieve a few core pancontigs.\\
